\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-21 papers/jss/article.Rnw
Revert Jereoen's two sentences about MessagePack and BSON back to my original three sentences, and add back the citations to the relevant R packages.
Jereon's version is very dismissive of these formats as not being
widely used or compatible with the almighty JSON.

My version notes that these address real deficiencies of JSON for the
application domain we are talking about in this paper, and points out
the shortcomings they still have compared to protocol buffers.

This introduction is a natural flow of alternatives each slightly
better than the last one discussed: starting with CSV, then XML

%TODO(de) 'protocol buffers' or 'Protocol Buffers' ?

\section{Introduction} % TODO(DE) More sober: Friends don't let friends use CSV}

Modern data collection and analysis pipelines increasingly involve collections
of decoupled components in order to better manage software complexity 
through reusability, modularity, and fault isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
% This is really a different pattern not connected well here.
%Data analysis patterns such as Split-Apply-Combine
%\citep{wickham2011split} explicitly break up large problems into manageable pieces.  
These pipelines are frequently built using different programming 
languages for the different phases of data analysis -- collection,
cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments and languages.  Each stage of such a data
analysis pipeline may produce intermediate results that need to be
stored in a file or sent over the network for further processing. 
% JO Perhaps also mention that serialization is needed for distributed
% systems to make systems scale up?
% MS: yes perhaps somewhere near here we could define serialization
% and describe this.

@@

%\paragraph*{Friends don't let friends use CSV!}
Data analysts and researchers often use character seperated text formats such
as \texttt{CSV} \citep{shafranovich2005common} to export and import
data. However, anyone who has ever used \texttt{CSV} files will have noticed
that this method has many limitations: it is restricted to tabular data,
lacks type-safety, and has limited precision for numeric values.  Moreover,
ambiguities in the format itself frequently cause problems.  For example,
conventions on which characters is used as seperator or decimal point vary by
country.  \emph{Extensible Markup Language} (\texttt{XML}) is another
well-established and widely-supported format with the ability to define just
about any arbitrarily complex schema \citep{nolan2013xml}. However, it pays
for this complexity with comparatively large and verbose messages, and added
complexitiy at the parsing side (which are somewhat mitigated by the
availability of mature libraries and parsers). Because \texttt{XML} is text
based and has no native notion of numeric types or arrays, it usually not a
very practical format to store numeric datasets as they appear in statistical
applications.  A more modern, widely used format is \emph{JavaScript Object
  Notation} (\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and used increasingly on the world wide web. \texttt{JSON} natively
supports arrays and distinguishes 4 primitive types: numbers, strings,
booleans and null. However, as it too is a text-based format, numbers are
stored as human-readable decimal notation which is somewhat inefficient and
leads to loss of type (double versus integer) and precision. Several R packages
implement functions to parse and generate \texttt{JSON} data from R objects.
A number of \texttt{JSON} variants has been proposed, such as \texttt{BSON}
and \texttt{MessagePack} which both add binary support. However, these
derivatives are not compatible with existing JSON software, and have not seen
wide adoption.

>>>>

%TODO(de) 'protocol buffers' or 'Protocol Buffers' ?
% MS: Lets standardize on 'Protocol Buffers'?

\section{Introduction} % TODO(DE) More sober: Friends don't let friends use CSV}
% NOTE(MS): I really do think we can use add back:
% \section{Introduction: Friends Don't Let Friends Use CSV}
% I didn't use proper Title Caps the first time around but really I
% think it makes the paper more readable to have a tl;dr intro title
% that is fun and engaging since this paper is still on the dry/boring
% side.
Modern data collection and analysis pipelines increasingly involve collections
of decoupled components in order to better manage software complexity 
through reusability, modularity, and fault isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
% This is really a different pattern not connected well here.
%Data analysis patterns such as Split-Apply-Combine
%\citep{wickham2011split} explicitly break up large problems into manageable pieces.  
These pipelines are frequently built using different programming 
languages for the different phases of data analysis -- collection,
cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments and languages.  Each stage of such a data
analysis pipeline may produce intermediate results that need to be
stored in a file or sent over the network for further processing. 
% JO Perhaps also mention that serialization is needed for distributed
% systems to make systems scale up?
% MS: yes perhaps somewhere near here we could define serialization
% and describe this.

@@

%\paragraph*{Friends don't let friends use CSV!}
Data analysts and researchers often use character seperated text formats such
as \texttt{CSV} \citep{shafranovich2005common} to export and import
data. However, anyone who has ever used \texttt{CSV} files will have noticed
that this method has many limitations: it is restricted to tabular data,
lacks type-safety, and has limited precision for numeric values.  Moreover,
ambiguities in the format itself frequently cause problems.  For example,
conventions on which characters is used as seperator or decimal point vary by
country.  \emph{Extensible Markup Language} (\texttt{XML}) is another
well-established and widely-supported format with the ability to define just
about any arbitrarily complex schema \citep{nolan2013xml}. However, it pays
for this complexity with comparatively large and verbose messages, and added
complexitiy at the parsing side (which are somewhat mitigated by the
availability of mature libraries and parsers). Because \texttt{XML} is text
based and has no native notion of numeric types or arrays, it usually not a
very practical format to store numeric datasets as they appear in statistical
applications.
%
A more modern, widely used format is \emph{JavaScript Object
  Notation} (\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and used increasingly on the world wide web. \texttt{JSON} natively
supports arrays and distinguishes 4 primitive types: numbers, strings,
booleans and null. However, as it too is a text-based format, numbers are
stored as human-readable decimal notation which is inefficient and
leads to loss of type (double versus integer) and precision. Several R packages
implement functions to parse and generate \texttt{JSON} data from R objects.

A number of binary formats based on \texttt{JSON} have been proposed
that reduce the parsing cost and improve efficiency.  \pkg{MessagePack}
\citep{msgpackR} and \pkg{BSON} \citep{rmongodb} both have R
interfaces, but these formats lack a separate schema for the seralized
data and thus still duplicate field names with each message sent over
the network or stored in a file.  Such formats also lack support for
versioning when data storage needs evolve over time, or when
application logic and requirement changes dictate updates to the
message format.

%and \texttt{MessagePack} which both add binary support. However, these
%derivatives are not compatible with existing JSON software, and have not seen
%wide adoption.
